---
title: "Reproducible Research Assignment"
author: "Carlos Ramirez"
date: "June 10, 2017"
output: html_document
---

This report shows the analysis and processing of the data about steps activity by time intervals recorded during the day recorded in the period from october to november, 2012.

### **1. Reading in the dataset and/or processing the data**

```{r}
## reading the data
activity <- read.csv(unz("activity.zip","activity.csv"),
                     header = TRUE,
                     colClasses = c("numeric","character","numeric"))
activity <- transform(activity, date = as.Date(date))
```

### **2. Total number of steps taken each day**

The next plot shows the distribution of dairy steps in all the data. 


```{r}
## calculating the total of steps
steps <- with(activity,tapply(steps,date,sum))

## plotting results
hist(steps,breaks = 50,col = "steelblue",
     main = "Number of dairy steps distribution",
     xlab = "Number of steps")
```

As can be seen the data is sparce but it can be seen that the more frequent number of dairy steps is around one thousand steps.

### **3. Mean and median of steps for each day**

The next plot shows the mean and median of the number of steps. Means (medians) are shown as red (green) points with respect to left (right) axis. 


```{r}
par( mar= c(5,5,5,5))
dates <- unique(activity$date)
means <- with(activity, tapply(steps,date, function(x) mean(x,na.rm = TRUE)))
medians <- with(activity, tapply(steps,date, function(x) median(x,na.rm = TRUE)))
plot(dates,means,pch=19,col="red",ylab = "Mean")
par(new=TRUE)
plot(dates,medians,pch=20,col="green",axes = FALSE,xlab = NA,ylab = NA)
axis(side=4)
mtext(side = 4,line = 3,"Median")
```

From the plot it can be seen that the median are not informative since in the vast majority of time intervals there are zero steps.

### **4. Time series plot of the average number of steps taken**

The next plot shows the time series of the average of steps over time intervals.

```{r}
## calculating average mean for intervals
intervals <- unique(activity$interval)
mean.av <- with(activity,tapply(steps, interval,function(x) mean(x,na.rm = TRUE)))

## plotting data
plot(intervals,mean.av,type = "l",col="red",
     main = "Average steps across day intervals",
     ylab = "Average mean",
     xlab = "Time intervals")
```

The plot shows that the more active intervals regarding number of steps is around the ~8.5 hours.

### **5. The 5-minute interval that, on average, contains the maximum number of steps**


```{r}
maxActivity <- intervals[which(mean.av==max(mean.av))]
maxActivity
```

The maximum average number of steps is then `r maxActivity` as is calculated in the above piece of code.

### **6. Code to describe and show a strategy for inputing missing data**


#### **6.1. Calculate and report the total number of missing values in the dataset**

The last entry of the summary applied to the steps column in the data shows the number of NA values as is shown next.

```{r}
with(activity, summary(steps))
```

The summary shows that there are 2304 NA values in the number of steps data.


#### **6.2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc**

In this report a simple strategy was used for inputing the missing data. Missing values were replaced by the mean number of steps of the corresponding day.

#### **6.3. Create a new dataset that is equal to the original dataset but with the missing data filled in**

The inputing method is implemented in the following code.

```{r}
## method for inputing the data
activity.inputed <- activity
for (i in 1:length(activity$steps)) {
        if ( is.na(activity$steps[i]) ) {
                day <- activity$date[i]
                activity.inputed$steps[i] <- mean(means,na.rm = TRUE)
        }
}
```

The activity.inputed data created in this manner have no missing number steps data as can be shown by using summary().

```{r}
with(activity.inputed, summary(steps))
```


#### **6.4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. **

The next code shows the analogous hystogram of the first figure in section 2 but using the inputed data.

```{r}
## caluculating total number of steps by day
wDays <- weekdays(activity.inputed$date)
activity.inputed$"weekDays" <- wDays
stepsByDay <- tapply(activity.inputed$steps,activity.inputed$date, function(x)                                                                    sum(x,na.rm = TRUE))

## plotting results
hist(stepsByDay,breaks = 30,col = "steelblue",
     main = "Number of steps (Inputed data)",
     xlab = "Number of steps")
```

No obvious qualitative change is seen in the inputed data in the distribution of the number of steps of each day. In the next code mean and median is caluculated for each day and plotted.

```{r}
## calculating the mean and median for each day
means.in <- tapply(activity.inputed$steps,activity.inputed$date, function(x)                                                                    mean(x,na.rm = TRUE))
medians.in <- tapply(activity.inputed$steps,activity.inputed$date, function(x)                                                                    median(x,na.rm = TRUE))

## plotting means and medians of inputed data
par(mar=c(5,5,5,5))
plot(dates,means.in,
     pch=19,col="red",ylab = "Mean")
par(new=TRUE)
plot(dates, medians.in,
     pch=20, col="green", 
     axes = FALSE, ylim = c(-1,max(means.in)+1), 
     xlab = NA, ylab = NA)
axis(side=4)
mtext(side = 4,line = 3,"Median")
```

From the plot it can be seen that medians are still no informative because for each day the majority of the values for intervals is zero steps. Means values now are filled with
mean values. 

#### **7. Histogram of the total number of steps taken each day after missing values are imputed**

```{r}
steps <- with(activity.inputed,tapply(steps,date,sum))
hist(steps,breaks = 50,col = "steelblue",
     main = "Number of dairy steps distribution (inputed data)",
     xlab = "Number of steps")
```

### **8. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends**

```{r}
## change days to weekday/weekend tags
activityByWeek <- activity.inputed
wDays <- weekdays(activity.inputed$date)
for (i in 1:length(wDays)) {
        if ( wDays[i] %in% c("sábado","domingo")) {
                wDays[i] <- "weekend"
        } else {
                wDays[i] <- "weekday"
        }
}
activity.inputed$weekDays <- wDays

## calculating the average of steps
activityByWeek <- split(activity.inputed,activity.inputed$weekDays)
activityByWeek <- lapply(activityByWeek,function(x) tapply(x$steps,x$interval, mean))

## plotting the results
par(mfrow=c(1,2))
par(mar=c(5,5,5,5))
plot(intervals,activityByWeek$weekday,
     type = "l",
     col="black",
     main = "WeekDay",
     xlab = "Time intervals",
     ylab = "Average steps")
par(mar=c(5,5,5,5))
plot(intervals,activityByWeek$weekend,
     type = "l",
     col="black",
     main = "Weekend",
     xlab = "",
     ylab = "")

```

From the plot above it can be seen that the studied subject is the more active in terms
the average of steps.

### **9. All of the R code needed to reproduce the results (numbers, plots, etc.) in the report**

All the scripts are given in the ... files included in the ... github project.

